Measuring causal influence with\\ back-to-back regression: the linear case

### 2019-05-23 paper/introduction.tex
typos

First, causal and non-causal factors may be numerous and collinear. In physics, for example, one may be challenged to identify whether fusion is caused by a change in temperature or a change in pressure, as these two factors may, at first, be difficult to manipulate independently. This issue becomes increasingly pronounced as the number of potential factor increases. In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity because word frequency covaries with other factors such as their length (short words are more frequent than long words), their categories (determinants are more frequent than adverbs), their individual letters (e.g. words that contain "z" are rare) and so forth \citep{kutas2011thirty,pegado2014timing}. Generating a set of words that control for all of these factors simultaneously can be challenging; it is thus common to use forward modeling, i.e. to train a model to predict observations (e.g. brain activity) from a minimal combination of competing factors (e.g. word length, word frequency, e.g. \citep{huth2016natural}), and investigate, in the model, the estimated contribution of each factor \citep{friston1994statistical}.

@@

Both forward and backward modeling have competing benefits and drawbacks. Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations. By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart2018deconstructing, king2018encoding}. There also exist cross-decomposition techniques such as Partial Least Squares (PLS) and Canonical Correlation Analysis (CCA), but they are notoriously difficult to interpretat \citep{lebart1995statistique}.

We introduce the 'back-to-back regression', which combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}) After detailing the method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques, both in terms of estimating causal influence and out-of-environment predictions (Section~\ref{sec:experiment_synthetic}). Finally, we apply back-to-back regression to a large neuroimaging dataset and reveal that distinct but collinear word features leads to distinguishable brain responses (Section~\ref{sec:experiment_real}).

>>>>

First, causal and non-causal factors may be numerous and collinear. In physics, for example, one may be challenged to identify whether fusion is caused by a change in temperature or a change in pressure, as these two factors may, at first, be difficult to manipulate independently. This issue becomes increasingly pronounced as the number of potential factors increases. In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity because word frequency covaries with other factors such as their length (short words are more frequent than long words), their categories (determinants are more frequent than adverbs), their individual letters (e.g. words that contain "z" are rare) and so forth \citep{kutas2011thirty,pegado2014timing}. Generating a set of words that control for all of these factors simultaneously can be challenging; it is thus common to use forward modeling, i.e. to train a model to predict observations (e.g. brain activity) from a minimal combination of competing factors (e.g. word length, word frequency, e.g. \citep{huth2016natural}), and investigate, in the model, the estimated contribution of each factor \citep{friston1994statistical}.

@@

Both forward and backward modeling have competing benefits and drawbacks. Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations. By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart2018deconstructing, king2018encoding}. There also exist cross-decomposition techniques such as Partial Least Squares (PLS) and Canonical Correlation Analysis (CCA), but they are notoriously difficult to interpret \citep{lebart1995statistique}.

We introduce the 'back-to-back regression', which combines the benefits of forward and backward modeling (Section~\ref{sec:algorithm}). After detailing the method and proving its convergence (Section~\ref{sec:theorem}), we show with synthetic data that it outperforms state-of-the-art forward, backward and cross-decomposition techniques, both in terms of estimating i) causal influence and ii) out-of-environment predictions (Section~\ref{sec:experiment_synthetic}). Finally, we apply back-to-back regression to a large neuroimaging dataset and reveal that distinct but collinear word features lead to distinguishable brain responses (Section~\ref{sec:experiment_real}).

### 2019-05-23 paper/paper.tex
typos

Back-to-Back Regression (B2B) consists in two steps.
%
First, we estimate the linear regression coefficients $\hat G$ from $Y$ to $X$, and construct the predictions $\hat X = Y \hat G$.
%
This backward regression recovers the correlations between $Y$ and each factor of $X$.
%
Second, we estimate the linear regression coefficients $\hat H$ from $X$ to $\hat X$.
%
The diagonal of the regression coefficients $\hat H$, denoted by $\hat{E} = \text{Diag}(E)$, is the desired estimate of the causal influence matrix $E$.

@@

The second regression is:
%
\begin{align*}
    \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
    &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]
\end{align*}
%
The solution to this second regression is $H = EF\hat{G}$.
%
Therefore, our estimate of $E$ is:
\begin{equation}
    \hat{E} = EF[(XEF)^\top (XEF) + (NF)^\top (NF)]^{-1} (XEF)^\top X.
    \label{eq:estimate}
\end{equation}
%
By the definition of the Moore-Penrose, if $(NF)^\top (NF)$ is full-rank with $\| (NF)^\top (NF) \| \to 0$, it follows that $\hat{E} = EF(XEF)^\dagger X$.
%
Next, assume that the true causal influence matrix contains $k$ ones arranged as $E = \left(\begin{array}{cc} I_k & 0 \\ 0 & 0 \end{array}\right)$, and let $E = K^\top K$, where $K\in \mathbb{R}^{k \times d_x}$.
%
    Also, let the covariance matrix of $X$ be $X^\top X = \left(\begin{array}{cc}\Sigma_{1} & \Sigma_{2} \\ \Sigma_{3} & \Sigma_{4}\end{array}\right)$, where $\Sigma_1 \in \mathbb{R}^{k\times k}$.
%
Then,
\begin{align*}
    \hat{E} &= EF(XEF)^\dagger X = K^\top K F(XK^\top K F)^\dagger X \stackrel{i)}{=} K^\top (K F) (KF)^\dagger (XK^\top)^\dagger X\\
            &\stackrel{ii)}{=} K^\top K(XK^\top)^\dagger X = E (XK^\top)^\dagger X \stackrel{iii)}{=} E (KX^\top X K^\top)^{-1} KX^\top X\\
            &= E (KX^\top X K^\top)^{-1} (\Sigma_1 \,|\, \Sigma_2)  = E \Sigma_1^{-1} (\Sigma_1 \,|\, \Sigma_2)
            = \left(\begin{array}{cc} I_k & \Sigma_1^{-1} \Sigma_2 \\ 0 & 0 \end{array}\right),
\end{align*}
where the equalities follow due to i) a full-rank factorization of $XEF$, ii) $KF$ being full-rank yields $(KF)(KF)^\dagger = K$, and iii) $X K^\top$ being full-rank yields $(XK^\top)^\dagger = (KX^\top X K^\top)^{-1} KX^\top$.
%
Therefore, $\text{Diag}(\hat{E})$ is the desired influence causal matrix $E$.
\end{proof}

>>>>

Back-to-Back Regression (B2B) consists of two steps.
%
First, we estimate the linear regression coefficients $\hat G$ from $Y$ to $X$, and construct the predictions $\hat X = Y \hat G$.
%
This backward regression recovers the correlations between $Y$ and each factor of $X$.
%
Second, we estimate the linear regression coefficients $\hat H$ from $X$ to $\hat X$.
%
The diagonal of the regression coefficients $\hat H$, denoted by $\hat{E} = \text{Diag}(E)$, is the desired estimate of the causal influence matrix $E$.

@@

The second regression is:
%
\begin{align*}
    \arg \min_H \mathbb{E}[\| XH - Y \hat{G} \|^2] &=\arg  \min_H \mathbb{E}[\| XH - (XE + N)F \hat G \|^2] \\
    &=\arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2] + \mathbb{E}[\| NF\hat G \| ^2]\\
    &= \arg \min_H \mathbb{E}[\| X(H - EF \hat G) \| ^2]
\end{align*}
%
The solution to this second regression is $H = EF\hat{G}$.
%
Therefore, our estimate of $E$ is:
\begin{equation}
    \hat{E} = EF[(XEF)^\top (XEF) + (NF)^\top (NF)]^{-1} (XEF)^\top X.
    \label{eq:estimate}
\end{equation}
%
By definition of the Moore-Penrose inverse (also known as pseudo-inverse), if $(NF)^\top (NF)$ is full-rank with $\|~(NF)^\top~(NF)~\|~\to~0$, it follows that $\hat{E} = EF(XEF)^\dagger X$.
%
Next, assume that the true causal influence matrix contains $k$ ones arranged as $E = \left(\begin{array}{cc} I_k & 0 \\ 0 & 0 \end{array}\right)$, and let $E = K^\top K$, where $K\in \mathbb{R}^{k \times d_x}$.
%
    Also, let the covariance matrix of $X$ be $X^\top X = \left(\begin{array}{cc}\Sigma_{1} & \Sigma_{2} \\ \Sigma_{3} & \Sigma_{4}\end{array}\right)$, where $\Sigma_1 \in \mathbb{R}^{k\times k}$.
%
Then,
\begin{align*}
    \hat{E} &= EF(XEF)^\dagger X = K^\top K F(XK^\top K F)^\dagger X \stackrel{i)}{=} K^\top (K F) (KF)^\dagger (XK^\top)^\dagger X\\
            &\stackrel{ii)}{=} K^\top K(XK^\top)^\dagger X = E (XK^\top)^\dagger X \stackrel{iii)}{=} E (KX^\top X K^\top)^{-1} KX^\top X\\
            &= E (KX^\top X K^\top)^{-1} (\Sigma_1 \,|\, \Sigma_2)  = E \Sigma_1^{-1} (\Sigma_1 \,|\, \Sigma_2)
            = \left(\begin{array}{cc} I_k & \Sigma_1^{-1} \Sigma_2 \\ 0 & 0 \end{array}\right),
\end{align*}
where the equalities follow due to i) a full-rank factorization of $XEF$, ii) $KF$ being full-rank yields $(KF)(KF)^\dagger = K$, and iii) $X K^\top$ being full-rank yields $(XK^\top)^\dagger = (KX^\top X K^\top)^{-1} KX^\top$.
%
Therefore, $\text{Diag}(\hat{E})$ is the desired causal influence matrix $E$.
\end{proof}
